### 2013-02-01 doc/xps2013/proposal.tex
partial commit for Luis the sick boy

Irregular applications generate tasks with work, interdependences, or memory
accesses that are highly sensitive to input. Classic examples of irregular
applications include branch and bound optimization, SPICE circuit simulation,
contact algorithms in car crash analysis, and network flow. Important
contemporary examples include processing large graphs in the business,
national security, machine learning, data-driven science, and social network
computing domains. For these emerging applications, fast response -- given the
sheer amount of data -- requires multinode systems. The most broadly available
multinode systems are those built from x86 commodity computing nodes
interconnected via ethernet or infiniband. Our goal is to enable scalable
performance of irregular applications on these mass market systems.

Our goal is challenging for two key reasons:

@@

During execution, a task yields control of its core whenever it performs
a long-latency operation, allowing the processor to remain busy while
waiting for the operation to complete.  In addition,  a programmer can
direct scheduling explicitly via the Grappa API calls shown
in Figure~\ref{fig:scheduling}.
To minimize yield overhead, the Grappa scheduler operates entirely in
user-space and does little more than store register state of one task and load
that of another. Context switch times are as low as 38ns even when switching
amongst thousands of tasks.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{ yield() } ] \hfill \\
      Yields core to scheduler, enqueuing caller to be scheduled again soon
    \item[ \texttt{ suspend() } ] \hfill \\
      Yields core to scheduler, enqueuing caller only once another task calls wake
    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
      Enqueues some other task $t$ to be scheduled again soon
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:scheduling} Grappa API: scheduling} %{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

@@

Grappa provides four methods for expressing parallelism, shown in Figure~\ref{fig:expressing-parallelism}:

First, when the programmer identifies work that can be done in parallel, the work may be wrapped up in a function and queued with its arguments for later execution using a \texttt{spawn}.

Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific core in the system or at the home core of a particular memory location.

Third, the programmer can invoke a parallel for loop, provided that the trip count is known at loop entry. The programmer specifies a function pointer along with start and end indices and an optional threshold to control parallel overhead. Grappa does {\em recursive decomposition} of iterations, similar to Cilk's cilk\_for construct~\cite {cilkforimplementation} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.  It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below the required threshold.

Fourth, a programmer may want to execute an active message; that is, to run a small piece of code on a particular core in the system without waiting for execution resources to be available.  For example, custom synchronization primitives execute this way, as a function executed on the core where the data lives.  Grappa provides the \texttt{call\_on} call for this purpose.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
      Creates a new stealable task
    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
      Creates a new private task that will run on a specific core 
    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
      Executes iterations of a loop as stealable tasks 
    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
      Runs a limited function on a specific core without consuming
      Grappa execution resources 
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

@@

Grappa also allows any local data on a core's stacks or heap to be
exported to the global address space to be made accessible to other
cores across the system. Addresses to global memory allocated in this
way use \textbf{2D global addresses}.  This uses a traditional PGAS
addressing model, where each address is a tuple of a rank in the job (or
global process ID) and an address in that process. The lower 48 bits of
the address hold a virtual address in the process. The top bit is set to
indicates that the reference is a 2D address (as opposed to linear
address). This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^{15}$ endpoints. Any node-local data can be made
accessible by other nodes in the system by wrapping the address and node
ID into a 2D global address. This address can then be accessed with a
delegate and can also be cached by other nodes. At the destination the
address is converted into a canonical x86 address by replacing the upper
bits with the sign-extended upper bit of the virtual address. 2D
addresses may refer to memory allocated from a single processes' heap or
from a task's stack. Figure~\ref{fig:memory-structure} shows how 2D and
linear addresses can refer to other cores' memory.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
      \item[ \texttt{ global\_address global\_malloc( size )} ] \hfill \\
      \item[ \texttt{ global\_free( global\_address )} ] \hfill \\
        Allocates and frees memory in the global heap
      \item[ \texttt{ delegate\_read( global\_address, local\_var )} ] 
      \item[ \texttt{ delegate\_write( global\_address, local\_var )} ] %\vspace{-2ex}
      \item[ \texttt{ delegate\_cas( global\_address, local\_var )} ] %\vspace{-2ex}
      \item[ \texttt{ delegate\_fetch\_inc( global\_address, local\_var )} ] %\vspace{-2ex} 
\hfill \\
        Performs a memory operation at the home core of a global address
      \item[ \texttt{ cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} ]
      \item[ \texttt{ cache\_release( global\_address, local\_buf )} ] %\vspace{-2ex} 
\hfill \\
Perform cache operations to acquire/release global data.  Acquire, returns a local pointer after all data has been copied to the local node.  Release, optionally writes data back to global memory and frees up management resources.
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:accessing-memory} Grappa API: accessing memory} %{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:memory-structure} Grappa memory structure}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

@@

\begin{figure*}[htb]
\begin{center}
  \includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
\begin{minipage}{1.9\columnwidth}
  \caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in
the global heap. It then spawns two tasks on remote cores to increment
elements of the array.  The first task increments two elements of the array using cache
operations. In step 2, the task is invoked. A cache request is issued
for two adjacent integers starting at the second element of the
array. Since these element are stored in the memories of two different
cores, this requires the sending of two messages in step 3. The task
is suspended until both responses arrive in step 4. The data carried
in these responses is stored in the local buffer. The elements are
then increment in the buffer. Then in step 5, the modified data is
sent back to the home node. Acknoledgements are returned in step 6 so
the task knows when the writes are complete. The second task increments an element of the array with a delegate
operation. In step 7, the task is invoked. A delegate request is sent
to the home core of the array element with the increment value. The
task suspends until the response is received. In step 8, the increment
is executed on the remote core. A response is returned in step 9 with
the previous value of the array element.
}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure*}

@@

%% UTS: performance comparison
\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.5\textwidth]{figs/uts_performance.pdf}
    \end{center}
    \caption{Performance of in-memory unbalanced tree search.}
    \label{fig:uts_compare}
\end{figure}

@@

\begin{figure}[tH]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/bfs_performance}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:bfs-performance} BFS performance}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\paragraph{BFS}

We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and
$2^{29}$ edges (25 GB of data). Figure~\ref{fig:bfs-performance} shows
our performance in terms of graph edges traversed per second. The XMT
is 2.5 times faster than Grappa at 16 nodes.  Performance does scale at a constant rate for Grappa, suggesting that adding more nodes will increase performance.

\begin{figure}[tH]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/centrality_performance}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:centrality-performance} Centrality performance}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\paragraph{Centrality}

We ran Betweenness Centrality on the same scale 25 Kronecker graph as
we did for BFS. Figure~\ref{fig:centrality-performance} shows our
performance in terms of graph edges traversed per second. At 16 XMT
processors/cluster nodes, the XMT is 1.75 times faster than Grappa.

>>>>

Irregular applications generate tasks with work, interdependences, or memory accesses that are highly sensitive to input. Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow. Important contemporary examples include processing large graphs in the business, national security, machine learning, data-driven science, and social network computing domains. For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems. The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband. Our goal is to enable scalable performance of irregular applications on these mass market systems.  Our goal is challenging for two key reasons:

@@

During execution, a task yields control of its core whenever it performs a long-latency operation, allowing the processor to remain busy while waiting for the operation to complete.  In addition,  a programmer can direct scheduling explicitly via the Grappa API calls. To minimize yield overhead, the Grappa scheduler operates entirely in user-space and does little more than store register state of one task and load that of another. Context switch times are as low as 38ns even when switching amongst thousands of tasks.

@@

Grappa provides four methods for expressing parallelism. First, when the programmer identifies work that can be done in parallel, the work may be wrapped up in a function and queued with its arguments for later execution using a \texttt{spawn}. Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific core in the system or at the home core of a particular memory location. Third, the programmer can invoke a parallel for loop, provided that the trip count is known at loop entry. The programmer specifies a function pointer along with start and end indices and an optional threshold to control parallel overhead. Grappa does {\em recursive decomposition} of iterations, similar to Cilk's cilk\_for construct~\cite {cilkforimplementation} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.  It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below the required threshold. Fourth, a programmer may want to execute an active message; that is, to run a small piece of code on a particular core in the system without waiting for execution resources to be available.  For example, custom synchronization primitives execute this way, as a function executed on the core where the data lives.

@@

Grappa also allows any local data on a core's stacks or heap to be exported to the global address space to be made accessible to other cores across the system. Addresses to global memory allocated in this way use \textbf{2D global addresses}.  This uses a traditional PGAS addressing model, where each address is a tuple of a rank in the job (or global process ID) and an address in that process. The lower 48 bits of the address hold a virtual address in the process. The top bit is set to indicates that the reference is a 2D address (as opposed to linear address). This leaves 15 bits for network endpoint ID, which limits our scalability to $2^{15}$ endpoints. Any node-local data can be made accessible by other nodes in the system by wrapping the address and node ID into a 2D global address. This address can then be accessed with a delegate and can also be cached by other nodes. At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address. 2D addresses may refer to memory allocated from a single processes' heap or from a task's stack.

@@

\begin{figure*}[htb]
\begin{center}
  \includegraphics[width=6in]{figs/delegate-cache}
\begin{minipage}{\columnwidth}

\caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in the global heap. It then spawns two tasks on remote cores to increment elements of the array.  The first task increments two elements of the array using cache operations. In step 2, the task is invoked. A cache request is issued for two adjacent integers starting at the second element of the array. Since these element are stored in the memories of two different cores, this requires the sending of two messages in step 3. The task is suspended until both responses arrive in step 4. The data carried in these responses is stored in the local buffer. The elements are then increment in the buffer. Then in step 5, the modified data is sent back to the home node. Acknoledgements are returned in step 6 so the task knows when the writes are complete. The second task increments an element of the array with a delegate operation. In step 7, the task is invoked. A delegate request is sent to the home core of the array element with the increment value. The task suspends until the response is received. In step 8, the increment is executed on the remote core. A response is returned in step 9 with the previous value of the array element. }

\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}

\hspace{-1in}\begin{minipage}{0.3\textwidth}
\begin{center}
\includegraphics[width=3in]{figs/uts_performance.pdf}
\caption{\label{fig:uts_compare} Performance of in-memory unbalanced tree search.}
\end{center}
\end{minipage}
\hspace{1.25in}\
\begin{minipage}{0.3\textwidth}
\begin{center}
\includegraphics[width=3in]{figs/bfs_performance}
\caption{\label{fig:bfs-performance} Performance of breadth-first searcch.}
\end{center}
\end{minipage}

\end{center}
\end{figure}

@@

\paragraph{BFS} We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and $2^{29}$ edges (25 GB of data). Figure~\ref{fig:bfs-performance} shows our performance in terms of graph edges traversed per second. The XMT is 2.5 times faster than Grappa at 16 nodes.  Performance does scale at a constant rate for Grappa, suggesting that adding more nodes will increase performance.

Our efforts are built on many existing ideas in programming languages, systems and architecture. In this section we discuss related frameworks and key enabling technologies that Grappa builds upon.

\paragraph{Comparable frameworks} Distributed graph processing frameworks like Pregel~\cite{pregel:2010} and Distributed GraphLab~\cite{distgraphlab:vldb12} share similar goals as Grappa. Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices. GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms.  GraphLab, however, imposes a rigid computation model where programmers must express computation as transformations on a vertex and its edge list only, with information only from adjacent vertexes. Pregel is only slightly less restrictive, as the input data can be any vertex in the graph.  Grappa also supports dynamic parallelism with asynchronous execution, but parallelism is expressed as tasks or loop iterations, which is a far more general programming model for irregular computation tasks.

\paragraph{Global memory} Grappa includes a custom implementation of a software distributed shared memory (DSM) system. Many traditional software DSM systems are page based~\cite{Treadmarks,munin} and aim to hide the fact that they are built in software from applications by exploiting the processor's paging mechanisms, therefore relying heavily on locality. Instead, Grappa, like other partitioned global address space (PGAS) models, implements its DSM at the language, rather than system level. Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and UPC~\cite{upc:2005} make accesses to shared structures look like normal memory references. As we describe later, Grappa chooses a middle ground, where global addresses are explicit in the API and local accesses are emitted conventionally by the compiler.  While Grappa's DSM system is conceptually similar to prior work, its implementation is tuned for irregular computations.  Past DSM work, being page-based, could exploit the RDMA capabilities of network hardware to move large page-sized blocks of data from node to node. In our experience, when these networks move small blocks of data (a few bytes), only a fraction of the available bandwidth is achieved. In addition, the DSM system in Grappa ends up being tightly coupled to the task scheduler in order to overlap long latency memory operations with useful computation.  For these two reasons it became necessary to build a new DSM system specifically for Grappa.

\paragraph{Multithreading} Grappa uses multithreading to tolerate memory latency. This is a well known technique. Hardware implementations include the Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife}, Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in this paper, Grappa use a lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks on a single processing core. The large number of tasks is required because of the extremely high internode latency Grappa is mitigating.  Grappa's task library employs several optimizations: an extremely fast task switch, a small task size, and judicious use of hardware prefetching to bring task state into the cache long before that task is actually scheduled.

\section{Proposed Research}

\subsection{Improved Networking layer}

\paragraph{Virtual topology:} Currently Grappa uses a flat network topology where each processing core in the system is individually addressed.  What this means, for example, is in a 128 node system, where each node is comprised of 16 processing cores, the network has 2048 end-points.  The aggregator treats each of these end-points individually.  This design choice is simple, and avoids lock contention within the underlying \gasnet layer, but does not scale.  As more system nodes are added, a quadratic increase in the number of threads is required in order to sustain aggregator performance.

We need to rewrite the aggregator to build in a hierarchical virtual network layer.  The first layer of this hierarchy is to treat all processors at a given system node as a single destination. We intend to continue to use the lock-free single-threaded \gasnet communication layer, as we have found the lock-based one to incur too high of an overhead for our purposes.  Thus the aggregator will need to very carefully utilize lock-free data-structures.

The second, and upper layers of the virtual topology are built from clustering system nodes into groups.  Instead of aggregating for all end-points, aggregation will occur for system nodes within a group.  A designated (and dedicated) node will act as a router, processing incoming and outgoing traffic to system nodes within the group and other groups.

Hierarchy will reduce scaling requirements from \BigO{n^2} threads to \BigO{n log(n)}.  Of course, it takes what used to be a single network hop for communication and turns it into \BigO{log(n)} hops.  More threads will be required to overcome this increased latency, but we believe only \BigO{log(n)} more, hence a real scaling benefit is achievable with the a virtual hierarchical network topology.

\paragraph{User-level RDMA:} Certain commodity networking hardware today supports remote direct memory access (RDMA).  Unfortunately, \emph{how} it is implemented is so crude that it is difficult or impossible to efficiently use in practice.  We start with a short description of the current semantics of RDMA implementations.  Next we describe the research we wish to undertake to make RDMA more useful.

Networking gear supports remote read and write operations.  Read operations take a source address and key (on a remote node), a destination address and key (on the local node), and a length.  Cards support a address translation for regions of memory that are registered with them.  These translations, indexed by the supplied key, are used to perform virtual to physical address translation.  The physical address of the destination address is similarly translated using the translation table at the local node.  Cards also support remote write operations with a similar set of semantics, where addresses are translated on cards using the cards local translation table.  Both read and write operations are asynchronous.  The requesting processor initiates the operation and then can check for completion by querying a completion queue at the local card.  \emph{Notably}, the processor(s) on the remote node are provided a polling interface whereby they can learn if a remote write operation occurred.  No such interface exists to learn about remote read or atomic operations.

This interface suffers from multiple shortcomings.  Ultimately, moving data to and from a remote node requires the interaction of two different operating system processes existing on two different system nodes.  Those processes each maintain a collection of runtime data structures and moving data among them means interacting with those data structures constructively.  Often such

\subsection{SSD-only data structures}

\subsection{Graph benchmark suite}
